---
title: "G4058 Final Project"
author: "Fu, Wen & Lau, Arnold"
date: "December 20, 2015"
output: html_document
---

####1. Introduction

This project will be using a dataset from Paralyzed Veterans of America (PVA) that was previously used for the 1998 Knowledge Discovery and Data Mining (KDD) Cup. The dataset consists of 191,779 donors who have donated to PVA at least once between June 1995 and June 1996, but who have not donated anything between June 1996 and June 1997. These are referred to as "lapsed" donors, or donors who made their last donation 13 to 24 months ago. These donors were subsequently all solicited for a promotional campaign in June 1997. The variable to be predicted is __TARGET\_B__, which tallies whether or not the donor responded to the June 1997 promotion.  

```{r}
train <- read.table("KDD/cup98lrn/cup98LRN.txt", header = TRUE, sep = ",", skipNul = TRUE)
dim(train)
test <- read.table("KDD/cup98val/cup98VAL.txt", header = TRUE, sep = ",", skipNul = TRUE)
# test_targets contains two response variables: whether or not a donor responded to the campaign, 
# and how much money they gave if they did. In this project we are only predicting whether or not
# a donor responded, i.e. classification.
test_targets <- read.table("KDD/valtargt.txt", header = TRUE, sep = ",", skipNul = TRUE)
library(dplyr)
# bind the response variable to the testing data via the unique id CONTROLN
test <- arrange(test, CONTROLN) 
test <- cbind(test, test_targets[,2:3])
rm(test_targets)
dim(test)
```

The dataset was previously divided by the contest organizers into a training set with 95,412 observations and a testing set with 96,367 observations. There are a total of 481 features in the dataset, some of which are more relevant than others. For this project we extracted only a selection of variables that we judged would be better predictors of the response variable. We also recoded some of the variables when appropriate. In order to make the recoding less redundant, we combine the training and testing data first and then resplit them after recoding is complete. Descriptions of each variable follow.  

```{r}
combi <- rbind(train, test)
sub <- c("PVASTATE", "DOMAIN", "AGE", "HOMEOWNR", "INCOME", "GENDER", "HIT", "MALEMILI",
         "MALEVET", "VIETVETS", "WWIIVETS", "LOCALGOV", "STATEGOV", "FEDGOV", "MAJOR", "WEALTH2", 
         "VETERANS", "PEPSTRFL", "ETH7", "ETH10", "ETH11", "AFC1", "AFC2", "AFC3", "AFC4", "AFC5", "AFC6",
         "VC1", "VC2", "VC3", "VC4", "NUMPRM12", "CARDGIFT", "TIMELAG", "AVGGIFT", "RFA_2R", "RFA_2F", 
         "RFA_2A", "RFA_3", "MDMAUD_R", "MDMAUD_F", "MDMAUD_A", "TARGET_B")
combi_sub <- combi[, sub]

# Recoding and cleaning

### PVASTATE_BIN is 1 if donor lives in a state served by a PVA chapter, 0 otherwise
combi_sub$PVASTATE_BIN[combi_sub$PVASTATE == "E" | combi_sub$PVASTATE == "P"] <- 1
combi_sub$PVASTATE_BIN[combi_sub$PVASTATE != "E" & combi_sub$PVASTATE != "P"] <- 0
combi_sub$PVASTATE_BIN <- as.factor(combi_sub$PVASTATE_BIN)
summary(combi_sub$PVASTATE_BIN)
```

```{r}
### URBANICITY --> urban, city, suburban, town, or rural
combi_sub$URBANICITY[substr(combi_sub$DOMAIN, 1, 1) == "U"] <- "Urban"
combi_sub$URBANICITY[substr(combi_sub$DOMAIN, 1, 1) == "C"] <- "City"
combi_sub$URBANICITY[substr(combi_sub$DOMAIN, 1, 1) == "S"] <- "Suburban"
combi_sub$URBANICITY[substr(combi_sub$DOMAIN, 1, 1) == "T"] <- "Town"
combi_sub$URBANICITY[substr(combi_sub$DOMAIN, 1, 1) == "R"] <- "Rural"
combi_sub$URBANICITY[combi_sub$DOMAIN == " "] <- "Not mentioned"
combi_sub$URBANICITY <- as.factor(combi_sub$URBANICITY)

### NBHOOD_SES --> highest, average, or lower. Urban areas included "above average" and "below average"
### categories, which we combine here into "average".

combi_sub$NBHOOD_SES[substr(combi_sub$DOMAIN, 2, 2) == "1"] <- "Highest SES"
combi_sub$NBHOOD_SES[substr(combi_sub$DOMAIN, 2, 2) == "2"] <- "Average SES" 
combi_sub$NBHOOD_SES[substr(combi_sub$DOMAIN, 2, 2) == "3" & 
                     substr(combi_sub$DOMAIN, 1, 1) == "U"] <- "Average SES"
combi_sub$NBHOOD_SES[substr(combi_sub$DOMAIN, 2, 2) == "3" &
                     substr(combi_sub$DOMAIN, 1, 1) != "U"] <- "Lowest SES"
combi_sub$NBHOOD_SES[substr(combi_sub$DOMAIN, 2, 2) == "4"] <- "Lowest SES"
combi_sub$NBHOOD_SES[combi_sub$DOMAIN == " "] <- "Not mentioned"

# HOMEOWNR_BIN is 1 if donor is a known homeowner and 0 if status is unknown or missing
combi_sub$HOMEOWNR_BIN[combi_sub$HOMEOWNR == "H"] <- 1
combi_sub$HOMEOWNR_BIN[combi_sub$HOMEOWNR == "U"] <- 0
combi_sub$HOMEOWNR_BIN[combi_sub$HOMEOWNR == " "] <- 0
combi_sub$HOMEOWNR_BIN <- as.factor(combi_sub$HOMEOWNR_BIN)

# Cleaning up GENDER
combi_sub$GENDER[combi_sub$GENDER == "A"] <- " "
combi_sub$GENDER[combi_sub$GENDER == "C"] <- " "
combi_sub$GENDER[combi_sub$GENDER == "U"] <- " "

### HIT has suspicious values "240" and "241" that don't look like they belong there

### Rename ETH7, ETH10 and ETH11 for descriptiveness
combi_sub$JAPANESE <- combi_sub$ETH7
combi_sub$KOREAN <- combi_sub$ETH10
combi_sub$VIETNAMESE <- combi_sub$ETH11

### I don't know how AFC2 differs from MALEMILI, how AFC5 differs from MALEVET, or how VC1 differs from VIETVETS
### or how VC3 differs from WWIIVETS

### Split RFA_3 into R, F, and A. RFA_3 describes promotions given out June 1996 (18 were given out April 1996)
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "F"] <- "First time donor"
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "I"] <- "Inactive donor"
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "N"] <- "New donor" 
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "A"] <- "Active donor"
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "L"] <- "Lapsing donor"
combi_sub$RFA_3R[substr(combi_sub$RFA_3, 1, 1) == "S"] <- "Star donor"
combi_sub$RFA_3R[combi_sub$RFA_3 == " "] <- "Not mailed"
combi_sub$RFA_3R <- as.factor(combi_sub$RFA_3R)

combi_sub$RFA_3F[substr(combi_sub$RFA_3, 2, 2) == "1"] <- "One gift"
combi_sub$RFA_3F[substr(combi_sub$RFA_3, 2, 2) == "2"] <- "Two gifts"
combi_sub$RFA_3F[substr(combi_sub$RFA_3, 2, 2) == "3"] <- "Three gifts"
combi_sub$RFA_3F[substr(combi_sub$RFA_3, 2, 2) == "4"] <- "Four or more gifts"
combi_sub$RFA_3F[combi_sub$RFA_3 == " "] <- "Not mailed"
combi_sub$RFA_3F <- as.factor(combi_sub$RFA_3F)

combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "A"] <- "$0.01 - $1.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "B"] <- "$2.00 - $2.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "C"] <- "$3.00 - $4.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "D"] <- "$5.00 - $9.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "E"] <- "$10.00 - $14.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "F"] <- "$15.00 - $24.99"
combi_sub$RFA_3A[substr(combi_sub$RFA_3, 3, 3) == "G"] <- "$25.00 and above"
combi_sub$RFA_3A[combi_sub$RFA_3 == " "] <- "Not mailed"
combi_sub$RFA_3A <- as.factor(combi_sub$RFA_3A)

train_sub <- combi_sub[1:95412, ]
test_sub <- combi_sub[95413:191779, ]

train_sub$TARGET_B <- as.factor(train_sub$TARGET_B)
test_sub$TARGET_B <- as.factor(test_sub$TARGET_B)
```

```{r}
glm1 <- glm(TARGET_B ~ RFA_3R + RFA_3F + RFA_3A + NUMPRM12 + CARDGIFT + AVGGIFT, data = train_sub, 
            family = "binomial"(link = "logit"))
glm1_pred <- predict(glm1, newdata = test_sub, type = "response")
cl_glm1_pred <- as.integer(glm1_pred > 0.5)
table(test_sub$TARGET_B, cl_glm1_pred)

library(rpart)
rpart1 <- rpart(TARGET_B ~ RFA_3R + RFA_3F + RFA_3A + NUMPRM12 + CARDGIFT + AVGGIFT + PEPSTRFL + MAJOR,
                data = train_sub, method = "class", control = rpart.control(cp = 0.0001))
rpart1_predict <- predict(rpart1, newdata = test_sub, type = "class")
```